---
title: "Using radiocarbon ages as a proxy for human population dynamics in late Pleistocene mainland Southeast Asia"
format: gfm
---

```{r}
library(tidyverse)
library(readxl)
library(rcarbon)

# Read excel sheet with dates across Southeast Asia and southern China
dates <- read_excel("Demographic analysis sites.xlsx")

# Calibrate dates using IntCal20
calDates <- calibrate(
  x = dates$C14Age,
  errors = dates$C14SD,
  calCurves = "intcal20",
  verbose = FALSE)

```

## Introduction


## Background


## Data

I gathered `r nrow(dates)` uncalibrated radiocarbon ages from 30 Pliestocene to early Holocene sites across Southeast Asia and southern China. For one of the Chinese sites Xiaodong I had to uncalibrate the ages using the built in feature in rcarbon. 

## Methods

First we calibrated all the dates using IntCal20 in rcarbon, then we made summed probability distributions and visualized it in multiple ways, first without binning. Next to account for well dated sites overshadowing sites with less amount of dates we binned the data by 100 years, then 200 years, then 500, and finally 5000. Another method used to account for inter-site variation is thinning where only 1 date from each bin is chosen to construct the SPD with. A new method used to plot was composite kernel density estimates where random calendar dates were sampled from each calibrated date to generate a kernel density estimate, its done multiple times which all are visualized as an envelope. After the initial SPD plots we decided to evaluate these our findings with a Monte-Carlo simulation approach to check if taphonomic loss and calibration process affected the shape of the plot. With an exponential growth model our Monte-Carlo simulations were used to check 5 timelines, Oldest Dryas to Bølling–Allerød Interstadial, Bølling–Allerød Interstadial to Younger Dryas, Marine Isotope 3 to Last Glacial Maximum, , 50,000 years ago to 8.2-kiloyear event and Last Glacial Maximum to 8.2-kiloyear event

## Results

The resulting plots are relatively consistent in peaks and show that practically every period is accounted for. The consistent increase in dates after 14,000 BP do call for some investigation which is our current focus.

```{r}
age_range <- c(50000,2000)

spd_unbinned <- spd(
  calDates,
  timeRange = age_range)
```

```{r}
#| label: fig-unbinned-climate
#| fig-cap: Unbinned SPD plot with climate events overlayed 


climate_events <- read_excel("Demographic analysis sites.xlsx",
                             sheet = 2) |> 
  mutate(midpoint = `End year` + ( `Start year` - `End year`) / 2 )


# with ggplot2
#| label: fig-unbinned-climate
#| fig-cap: Unbinned SPD plot with climate events overlayed 

climate_events <- read_excel("Demographic analysis sites.xlsx", sheet = 2) |>
  filter(!is.na(`Climate event`)) |>
  mutate(
    Abbreviation = as.character(Abbreviation),
    midpoint = `End year` + (`Start year` - `End year`) / 2,
    colour = case_when(
      Condition == "warmer" ~ "salmon",
      Condition == "cooler" ~ "skyblue",
      is.na(Condition)      ~ "grey50"
    ),
    ymax = case_when(
  Abbreviation %in% c("LGM", "OD", "BA", "YD", "8.2k") ~ 0.09,
  is.na(Condition) ~ Inf,
  row_number() %% 2 == 0 ~ 0.08,
  TRUE ~ 0.06
    ),
    y_text = case_when(
  Abbreviation %in% c("MIS 3", "MIS 2", "MIS 1") ~ 0.095,
  Abbreviation %in% c("LGM", "OD", "BA", "YD", "8.2k") ~ 0.08,
  is.na(Condition) ~ 0.09,  # Heinrich events
  row_number() %% 2 == 0 ~ 0.07,
  TRUE ~ 0.05
    )
  )

climate_layers <- lapply(seq_len(nrow(climate_events)), function(i) {
  e <- climate_events[i, ]
  if (e$`Start year` == e$`End year`) {
    # Point events like Heinrich events → vertical line
    list(
      annotate("segment",
               x = e$`Start year`, xend = e$`Start year`,
               y = 0, yend = Inf,
               colour = "grey50", linetype = "dashed"),
      annotate("text",
               x = e$midpoint, y = 0.09,
               label = e$Abbreviation)
    )
  } else {
    list(
      annotate("rect",
               ymin = 0, ymax = e$ymax,
               xmin = e$`Start year`, xmax = e$`End year`,
               colour = e$colour, fill = e$colour,
               alpha = if_else(is.na(e$Condition), 0.1, 0.2)),
      annotate("text",
               x = e$midpoint, y = e$y_text,
               label = e$Abbreviation)
    )
  }
})

# Define this once alongside climate_events and climate_layers
climate_layers_for <- function(start_bp, end_bp) {
  events <- climate_events |> 
    filter(`Start year` <= start_bp & `End year` >= end_bp)
  
  lapply(seq_len(nrow(events)), function(i) {
    e <- events[i, ]
    if (e$`Start year` == e$`End year`) {
      list(
        annotate("segment",
                 x = e$`Start year`, xend = e$`Start year`,
                 y = 0, yend = Inf,
                 colour = "grey50", linetype = "dashed"),
        annotate("text",
                 x = e$midpoint, y = 0.09,
                 label = e$Abbreviation)
      )
    } else {
      list(
        annotate("rect",
                 ymin = 0, ymax = Inf,
                 xmin = e$`Start year`, xmax = e$`End year`,
                 colour = e$colour, fill = e$colour,
                 alpha = if_else(is.na(e$Condition), 0.1, 0.2)),
        annotate("text",
                 x = e$midpoint, y = e$y_text,
                 label = e$Abbreviation)
      )
    }
  })
}


ggplot(spd_unbinned$grid) +
  aes(calBP, PrDens) +
  geom_line() +
  climate_layers+
  scale_x_reverse() +
  theme_minimal() 
```
In @fig-unbinned-climate there is a continuous occupation of Southeast Asia from 25000 BP to present although there is a more consistent increase in height starting from 14 ka.


```{r}
# Initial Monte Carlo test for OD to BA
expnull <- modelTest(calDates, 
                      errors = dates$C14SD, 
                      model = "exponential", 
                      timeRange = c(18000, 12900), 
                      verbose = FALSE,
                      nsim = 100)
```

```{r}
#| label: fig-MTCL1
#| fig-cap: Monte Carlo test of OD-BA with 100 simulations 

plot_df <- expnull$result
#Annotating climate events
ggplot(plot_df, aes(x = calBP)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), fill = "grey70", alpha = 0.8) +
  geom_line(aes(y = PrDens), colour = "black") +
  # Highlight periods where observed exceeds envelope
  geom_vline(
    data = subset(plot_df, PrDens > hi | PrDens < lo),
    aes(xintercept = calBP),
    colour = "green", alpha = 0.3, linewidth = 0.3
  ) +
  scale_x_reverse(
    limits = c(18000, 12900),
    breaks = seq(18000, 13000, by = -1000)
  ) +
  labs(x = "Years cal BP", y = "Summed Probability") +
  climate_layers_for(18000, 12900)+
  theme_classic()

```

@fig-MTCL1 shows our testing from the Oldest Dryas to Bølling–Allerød Interstadial, we see 2 minor deviations where the line goes above the envelope near 14.1ka and 13.1ka.

```{r}
# Initial Monte Carlo test for OD to BA with 1000 simulations
expnull1000 <- modelTest(calDates, 
                      errors = dates$C14SD, 
                      model = "exponential", 
                      timeRange = c(18000, 12900), 
                      verbose = FALSE,
                      nsim = 1000)
```

```{r}
#| label: fig-MTCL1000
#| fig-cap: Monte Carlo test of OD-BA with 1000 simulations 

plot_df1000 <- expnull1000$result
# Annotating climate events
ggplot(plot_df1000, aes(x = calBP)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), fill = "grey70", alpha = 0.8) +
  geom_line(aes(y = PrDens), colour = "black") +
  # Highlight periods where observed exceeds envelope
  geom_vline(
    data = subset(plot_df1000, PrDens > hi | PrDens < lo),
    aes(xintercept = calBP),
    colour = "green", alpha = 0.3, linewidth = 0.3
  ) +
  scale_x_reverse(
    limits = c(18000, 12900),
    breaks = seq(18000, 13000, by = -500)
  ) +
  labs(x = "Years cal BP", y = "Summed Probability") +
  climate_layers_for(18000, 12900)+
  theme_classic()

```

In @fig-MTCL1000 we increased the simulations to 1000, the 2 deviations still exist and the line falls in the expected envelope through the whole time.

```{r}
# Initial Monte Carlo test for BA-YD
expnull2 <- modelTest(calDates, 
                      errors = dates$C14SD, 
                      model = "exponential", 
                      timeRange = c(15000, 11000), 
                      verbose = FALSE,
                      nsim = 100)
```

```{r}
#| label: fig-MTCL2
#| fig-cap: Monte Carlo test of BA-YD with 100 simulations  

plot_df2 <- expnull2$result

# Annotating climate events
ggplot(plot_df2, aes(x = calBP)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), fill = "grey70", alpha = 0.8) +
  geom_line(aes(y = PrDens), colour = "black") +
  # Highlight periods where observed exceeds envelope
  geom_vline(
    data = subset(plot_df2, PrDens > hi | PrDens < lo),
    aes(xintercept = calBP),
    colour = "green", alpha = 0.3, linewidth = 0.3
  ) +
  scale_x_reverse(
    limits = c(15000, 11000),
    breaks = seq(15000, 11000, by = -500)
  ) +
  labs(x = "Years cal BP", y = "Summed Probability") +
  climate_layers_for(15000, 11000)+
  theme_classic()
```

@fig-MTCL2 shows our testing from Bølling–Allerød Interstadial to Younger Dryas, the same deviation near 13.1ka is seen again while there is a new one in the Younger Dryas near 12.9ka.

```{r}
# Initial Monte Carlo test for 50ka to LGM
expnull3 <- modelTest(calDates, 
                      errors = dates$C14SD, 
                      model = "exponential", 
                      timeRange = c(50000, 19000), 
                      verbose = FALSE,
                      nsim = 100)
```

```{r}
#| label: fig-MTCL3
#| fig-cap: Monte Carlo test of MIS3-LGM with 100 simulations 

plot_df3 <- expnull3$result
# Annotating climate events
ggplot(plot_df3, aes(x = calBP)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), fill = "grey70", alpha = 0.8) +
  geom_line(aes(y = PrDens), colour = "black") +
  # Highlight periods where observed exceeds envelope
  geom_vline(
    data = subset(plot_df3, PrDens > hi | PrDens < lo),
    aes(xintercept = calBP),
    colour = "green", alpha = 0.01, linewidth = 0.1
  ) +
  scale_x_reverse(
    limits = c(57000, 19000),
    breaks = seq(57000, 19000, by = -5000)
  ) +
  labs(x = "Years cal BP", y = "Summed Probability") +
  climate_layers_for(57000, 19000)+
  theme_classic()

```

@fig-MTCL3 shows our test from MIS3 to LGM, there is a positive deviation nearly the entire time up to the LGM

```{r}
# Initial Monte Carlo test for 50ka to 8.2
expnull4 <- modelTest(calDates, 
                      errors = dates$C14SD, 
                      model = "exponential", 
                      timeRange = c(50000, 8000), 
                      verbose = FALSE,
                      nsim = 100)
```

```{r}
#| label: fig-MTCL4
#| fig-cap: Monte Carlo test of 50K-8.2k with 100 simulations 

plot_df4 <- expnull4$result

# Annotating climate events
ggplot(plot_df4, aes(x = calBP)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), fill = "grey70", alpha = 0.8) +
  geom_line(aes(y = PrDens), colour = "black") +
  # Highlight periods where observed exceeds envelope
  geom_vline(
    data = subset(plot_df4, PrDens > hi),
    aes(xintercept = calBP),
    colour = "green", alpha = 0.01, linewidth = 0.1
  ) +
  geom_vline(
    data = subset(plot_df4, PrDens < lo),
    aes(xintercept = calBP),
    color = "blue", alpha = 0.01, linewidth = 0.1
  )+
  scale_x_reverse(
    limits = c(50000, 8000),
    breaks = seq(50000, 8000, by = -5000)
  ) +
  labs(x = "Years cal BP", y = "Summed Probability") +
  climate_layers_for(50000, 8000)+
  theme_classic()
```

@fig-MTCL4 shows our test from 50ka to the 8.2 kiloyear event, we can also see a near constant positive deviation up to the MIS2 then a few positive deviations during the LGM and then the line follows the envelope until the 8.2 kiloyear event where there is a negative deviation.


```{r}
# Initial Monte Carlo test for LGM to 8.2
expnull5 <- modelTest(calDates, 
                      errors = dates$C14SD, 
                      model = "exponential", 
                      timeRange = c(26500, 8000), 
                      verbose = FALSE,
                      nsim = 100)
```

```{r}
#| label: fig-MTCL5
#| fig-cap: Monte Carlo test of LGM-8.2k with 100 simulations 

plot_df5 <- expnull5$result

# Annotating climate events
ggplot(plot_df5, aes(x = calBP)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), fill = "grey70", alpha = 0.8) +
  geom_line(aes(y = PrDens), colour = "black") +
  # Highlight periods where observed exceeds envelope
  geom_vline(
    data = subset(plot_df5, PrDens > hi),
    aes(xintercept = calBP),
    colour = "green", alpha = 0.3, linewidth = 0.3
  ) +
  geom_vline(
    data = subset(plot_df4, PrDens < lo),
    aes(xintercept = calBP),
    color = "blue", alpha = 0.01, linewidth = 0.1
  )+
  scale_x_reverse(
    limits = c(26500, 8000),
    breaks = seq(26500, 8000, by = -5000)
  ) +
  labs(x = "Years cal BP", y = "Summed Probability") +
  climate_layers_for(26500, 8000)+
  theme_classic()
```

@fig-MTCL5 shows our test for LGM to the 8.2 kiloyear event there are a few minor positive deviations near 23ka, 24ka, 18ka, 16ka, 13ka, and a few minor negative deviations near the 8.2 kiloyear event.

```{r}
#| label: fig-200-bin
#| fig-cap: SPD plot with a bin of 200 years

# Grouping dates within 200 years of each other into bins
bins_200 <- binPrep(
  sites = dates$Site,
  ages = dates$C14Age,
  h = 200
)

# Plotting the summed probability distribution with new bins
spd_res <- spd(
  calDates, 
  bins = bins_200,
  timeRange = age_range,
  verbose = FALSE)
```

```{r}
#| label: fig-200-bin-climate
#| fig-cap: SPD plot with a bin of 200 years with climate events overlayed

ggplot(spd_res$grid) +
  aes(calBP, PrDens) +
  geom_line() +
  climate_layers+
  scale_x_reverse() +
  theme_minimal() 
```

In @fig-200-bin-climate the changes are more noticeable as the peak around 13 ka and just before 10 ka is much higher than the original plot.

```{r}
#| label: fig-map
#| fig-cap: "Map of early sites in Southeast Asia"

library(tidyverse)
library(readxl)
library(rnaturalearth)
library(rnaturalearthdata)
library(sf)
library(ggrepel)

# Load site data from Excel file
dates_data <- read_excel("Demographic analysis sites.xlsx")

# Prepare the spatial data by filtering unique sites
sites_sf <- dates_data %>%
  distinct(Site, lat, long) %>%
  st_as_sf(coords = c("long", "lat"), 
           remove = FALSE, 
           crs = 4326)

# Get the map background
world <- ne_countries(scale = "medium", returnclass = "sf")

# Generate the Map, focus on Southeast Asia, and label sites clearly 
sea_map <- ggplot(world) +
  geom_sf() +
  geom_sf(data = sites_sf, size = 2) +
  coord_sf(xlim = c(90, 115), 
           ylim = c(8, 30), 
           expand = FALSE) +
  geom_text_repel(data = sites_sf,
                  aes(x = long, y = lat, label = Site),
                  max.overlaps = Inf,
                  size = 3, 
                  nudge_x = 0.5,
                  force = 10,
                  point.padding = 0.5,
                  box.padding = 0.5,
                  min.segment.length = 0,
                  bg.color = "white",
                  bg.r = 0.15) +
  theme_void() +
  labs(title = "Early sites in Southeast Asia") 

sea_map
```
In @fig-map are all the sites in early Mainland Southeast Asia used in this study 


```{r}
#| label: fig-early-map
#| fig-cap: "Sites older than 30,000 BP"
# Filter for sites older than 30,000 BP
earliest_sites <- dates_data %>%
  filter(C14Age > 30000) %>%
  distinct(Site, lat, long) %>%
  st_as_sf(coords = c("long", "lat"), 
           remove = FALSE, 
           crs = 4326)

# Base map of Southeast Asia
world <- ne_countries(scale = "medium", returnclass = "sf")

# Plot earliest sites
early_sea_map <-ggplot(world) +
  geom_sf() +
  geom_sf(data = earliest_sites, size = 2) +
  coord_sf(xlim = c(90, 115), 
           ylim = c(8, 30), 
           expand = FALSE) +
  geom_text_repel(data = earliest_sites,
                  aes(x = long, y = lat, label = Site),
                  max.overlaps = Inf,
                  size = 3, 
                  nudge_x = 0.5,
                  force = 10,
                  point.padding = 0.5,
                  box.padding = 0.5,
                  min.segment.length = 0,
                  bg.color = "white",
                  bg.r = 0.15) +
  theme_void() +
  labs(title = "Sites older than 30 ka") 

early_sea_map
```
In @fig-early-map the oldest sites from before 30,000 are shown

```{r}
#| label: fig-after-map
#| fig-cap: "Sites occupied after the 14 ka population rise"
# Filter for sites after the 14 ka event
sites_after_14ka <- dates_data %>%
  filter(C14Age <= 13000) %>%
  distinct(Site, lat, long) %>%
  st_as_sf(coords = c("long", "lat"), 
           remove = FALSE, 
           crs = 4326)

# Plot the after 14 ka sites
sea_map_after_14ka <-ggplot(world) +
  geom_sf() +
  geom_sf(data = sites_after_14ka, size = 2) +
  coord_sf(xlim = c(90, 115), 
           ylim = c(8, 30), 
           expand = FALSE) +
  geom_text_repel(data = sites_after_14ka,
                  aes(x = long, y = lat, label = Site),
                  max.overlaps = Inf,
                  size = 3, 
                  nudge_x = 0.5,
                  force = 10,
                  point.padding = 0.5,
                  box.padding = 0.5,
                  min.segment.length = 0,
                  bg.color = "white",
                  bg.r = 0.15) +
  theme_void() +
  labs(title = "Occupation after 13 ka") 

sea_map_after_14ka
```
In @fig-after-map sites occupied after 13,000 are shown, notably there is an increase in sites in southern Mainland Southeast Asia

```{r}
#| label: fig-karst-map
#| fig-cap: "Early sites with Karst layered on top"
# Import the karst data
karst_sf <- st_read("whymap_karst__v1_poly.shp")

ggplot(world) +
  geom_sf(fill = "white") +              # Base map
  geom_sf(data = karst_sf,                     # Karst data
          color = "steelblue1",
          fill = "steelblue1",
          alpha = 1, 
          size = 1) +
  geom_sf(data = sites_sf, size = 2) + # Existing sites 
  coord_sf(xlim = c(90, 115), 
           ylim = c(8, 30), 
           expand = FALSE) +
  geom_text_repel(data = sites_sf,
                  aes(x = long, y = lat, label = Site),
                  max.overlaps = Inf,
                  size = 3, 
                  nudge_x = 0.5,
                  force = 10,
                  point.padding = 0.5,
                  box.padding = 0.5,
                  min.segment.length = 0,
                  bg.color = "white",
                  bg.r = 0.15) +
  theme_minimal() +
  labs(title = "Archaeological Sites and Karst Formations")
```
In @fig-karst-map all sites are shown with karst formations overlayed


