---
title: "SPD test"
format: gfm
---

## Data
I gathered 269 uncalibrated radiocarbon ages from 32 Pliestocene to early Holocene sites across Southeast Asia and southern China. For one of the Chinese sites Xiaodong I had to uncalibrate the ages using the built in feature in rcarbon. 

## Methods
First we calibrated all the dates using IntCal20 in rcarbon, then we made summed probability distributions and visualized it in multiple ways, first without binning. Next to account for well dated sites overshadowing sites with less amount of dates we binned the data by 100 years, then 200 years, then 500, and finally 5000. Another method used to account for inter-site variation is thinning where only 1 date from each bin is chosen to construct the SPD with. A new method used to plot was composite kernel density estimates where random calendar dates were sampled from each calibrated date to generate a kernel density estimate, its done multiple times which all are visualized as an envelope. 
```{r}
library(readxl)
library(rcarbon)

# Read excel sheet with dates across Southeast Asia and southern China
dates <- read_excel("Demographic analysis sites.xlsx")

# Calibrate dates using IntCal20
calDates <- calibrate(
  x = dates$C14Age,
  errors = dates$C14SD,
  calCurves = "intcal20"  
)
```

## Results
The resulting plots are relatively consistent in peaks and show that practically every period is accounted for. The consistent increase in dates after 14,000 ka do call for some investigation which is our current focus.
```{r}
# Plotting the summed probability distribution
spd_unbinned <- spd(
  calDates,
  timeRange = c(25000, 5000))
plot(spd_unbinned)
```

```{r}
# Grouping dates within 100 years of each other into bins
bins_100 <- binPrep(
  sites = dates$Site,
  ages = dates$C14Age,
  h = 100
)

# Plotting the summed probability distribution with new bins
spd_res <- spd(
  calDates, 
  bins = bins_100,
  timeRange = c(25000, 5000))
plot(spd_res)
```


```{r}
# Grouping dates within 200 years of each other into bins
bins_200 <- binPrep(
  sites = dates$Site,
  ages = dates$C14Age,
  h = 200
)

# Plotting the summed probability distribution with new bins
spd_res <- spd(
  calDates, 
  bins = bins_200,
  timeRange = c(25000, 5000))
plot(spd_res)
```


```{r}
# Grouping dates within 500 years of each other into bins
bins_500 <- binPrep(
  sites = dates$Site,
  ages = dates$C14Age,
  h = 500
)

# Plotting the summed probability distribution with new bins
spd_res <- spd(
  calDates, 
  bins = bins_500,
  timeRange = c(25000, 5000))
plot(spd_res)
```

```{r}
# Grouping dates within 5000 years of each other into bins
bins_5000 <- binPrep(
  sites = dates$Site,
  ages = dates$C14Age,
  h = 5000
)

# Plotting the summed probability distribution with new bins
spd_res <- spd(
  calDates, 
  bins = bins_5000,
  timeRange = c(25000, 5000))
plot(spd_res)
```

```{r}
# Grouping dates within 200 years of each other into bins
bins_200 <- binPrep(
  sites = dates$Site,
  ages = dates$C14Age,
  h = 200
)

# Thinning selects 1 random date from each bin
calDates2 = calDates[thinDates(ages = dates$C14Age,
                               errors = dates$C14SD,
                               bins = bins_200,
                               size = 1,
                               method = 'random')]

spd_thin <- spd(
  calDates2,
  timeRange = c(25000, 5000)
)

# Plotting the SPD after thinning
plot(spd_thin)
```

```{r}
# Grouping dates within 200 years of each other into bins
bins_200 <- binPrep(
  sites = dates$Site,
  ages = dates$C14Age,
  h = 200
)

# Randomly sampling from each calibrated date to generate a kernel density estimate
ckde_res = sampleDates(calDates,bins=bins_200,nsim=100,verbose=FALSE)

Sea.ckde = ckde(ckde_res,
                timeRange = c(25000, 5000),
                bw = 200)
#Plotting the CKDE
plot(Sea.ckde, type = 'multiline')
```



